---
title: "Sequential Poisson Sampling"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

[![CRAN status](https://www.r-pkg.org/badges/version/sps)](https://cran.r-project.org/package=sps)
[![R-CMD-check](https://github.com/marberts/sps/workflows/R-CMD-check/badge.svg)](https://github.com/marberts/sps/actions)
[![codecov](https://codecov.io/gh/marberts/sps/branch/master/graph/badge.svg?token=5CPGWUF267)](https://app.codecov.io/gh/marberts/sps)

Sequential Poisson sampling is a method for drawing probability-proportional-to-size samples with a given number of units, and is commonly used for price-index surveys. This package gives functions to draw a stratified sequential Poisson sample according to the method in Ohlsson (1998), and generate bootstrap replicate weights according to the method in Beaumont and Patak (2012).

## Installation

```r
install.packages("sps")
```

The latest version can be found on GitHub.

```r
devtools::install_github("marberts/sps")
```

## Usage

Given a vector of sizes for units in a population (e.g., revenue for sampling businesses) and a desired sample size, a stratified sequential Poisson sample can be drawn with the `sps()` function. 

```{r}
library(sps)

# Generate some data on sizes for 12 businesses in a single 
# stratum as a simple example
revenue <- c(1:10, 100, 150)

# Draw a sample of 6 businesses
(samp <- sps(revenue, 6))

# Design weights and sampling strata are stored with the sample
weights(samp)
levels(samp)
```

The design weights for a sample can then be used to generate bootstrap replicate weights with the `sps_repwights()` function.

## References

Beaumont, J.-F., and Patak, Z. (2012). On the Generalized Bootstrap for Sample Surveys with Special Attention to Poisson Sampling. *International Statistical Review*, 80(1): 127-148.

Ohlsson, E. (1998). Sequential Poisson Sampling. *Journal of Official Statistics*, 14(2): 149-162.