---
title: "Sequential Poisson Sampling"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

[![CRAN status](https://www.r-pkg.org/badges/version/sps)](https://cran.r-project.org/package=sps)
[![sps status badge](https://marberts.r-universe.dev/badges/sps)](https://marberts.r-universe.dev)
[![R-CMD-check](https://github.com/marberts/sps/workflows/R-CMD-check/badge.svg)](https://github.com/marberts/sps/actions)
[![codecov](https://codecov.io/gh/marberts/sps/branch/master/graph/badge.svg?token=5CPGWUF267)](https://app.codecov.io/gh/marberts/sps)

Sequential Poisson sampling is a method for drawing probability-proportional-to-size samples with a given number of units, and is commonly used for price-index surveys. This package gives functions to draw stratified sequential Poisson samples according to the method by Ohlsson (1998), and generate bootstrap replicate weights according to the method by Beaumont and Patak (2012).

## Installation

```r
install.packages("sps")
```

The development version can be found on GitHub.

```r
devtools::install_github("marberts/sps")
```

## Usage

Given a vector of sizes for units in a population (e.g., revenue for sampling businesses) and a desired sample size, a stratified sequential Poisson sample can be drawn with the `sps()` function. 

```{r}
library(sps)

# Generate some data on sizes for 12 businesses in a single 
# stratum as a simple example
revenue <- c(1:10, 100, 150)

# Draw a sample of 6 businesses
(samp <- sps(revenue, 6))

# Design weights and sampling strata are stored with the sample
weights(samp)
levels(samp)
```

Allocations are often proportional to size when drawing such samples, and the `prop_allocation()` function provides a variety of methods for generating proportional-to-size allocations.

```{r}
# Add some strata
stratum <- rep(c("a", "b"), c(9, 3))

# Make an allocation
(allocation <- prop_allocation(revenue, 6, stratum))

# Draw a stratified sample
(samp <- sps(revenue, allocation, stratum))

weights(samp)
levels(samp)
```

The design weights for a sample can then be used to generate bootstrap replicate weights with the `sps_repwights()` function.

```{r}
sps_repweights(weights(samp), 5, tau = 2)
```

## References

Beaumont, J.-F. and Patak, Z. (2012). On the Generalized Bootstrap for Sample Surveys with Special Attention to Poisson Sampling. *International Statistical Review*, 80(1): 127-148.

Ohlsson, E. (1998). Sequential Poisson Sampling. *Journal of Official Statistics*, 14(2): 149-162.